Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:ct}
 \begin{align*}
  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
                                 & G(x) \wedge W(x) \wedge T_o(x, x') \rightarrow W(x') \wedge R(x) > 0 
  \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

At this point, we know how to construct two formulae for a loop $L$: one
that is satisfiable iff $L$ is terminating and another that is satisfiable
iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$,
respectively:
%
\begin{align*}
 \exists P_N . \forall x, x' . \phi(W_N, x, x') \\
 \exists P_T . \forall x, x' . \psi(W_T, x, x')
\end{align*}
%
We can combine these:
%
\begin{align*}
 (\exists P_N . \forall x, x'. \phi(W_N, x, x')) \vee (\exists P_T . \forall x, x', \psi(W_T, x, x'))
\end{align*}
%
Which simplifies to:
%
\begin{definition}[Generalised Termination Formula]
\begin{align*}
 \exists P_N, P_T. \forall x, x', y, y', \phi(P_N, x, x') \vee \psi(P_T, y, y') 
\end{align*}
\end{definition}

Since $L$ either terminates or does not terminate, this formula is a
tautology in the synthesis fragment.  A solution to the formula would
include witnesses $P_N$ and $P_T$, which are putative proofs of
non-termination and termination, respectively.  Exactly one of these will be
a genuine proof, so we can check first one and then the other.

>>>>

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:ct}
 \begin{align*}
  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
                                 & G(x) \wedge W(x) \wedge T(x, x') \rightarrow W(x') \wedge R(x) > 0 
  \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

At this point, we know how to construct two formulae for a loop $L$: one
that is satisfiable iff $L$ is terminating and another that is satisfiable
iff it is non-terminating.  We will call these formulae $\phi$ and $\psi$,
respectively:
%
\begin{align*}
 \exists P_T . \forall x, x' . \phi(P_T, x, x') \\
 \exists P_N . \forall x . \psi(P_N, x)
\end{align*}
%
We can combine these:
%
\begin{align*}
 (\exists P_T . \forall x, x'. \phi(P_T, x, x')) \vee (\exists P_N . \forall x, \psi(P_N, x))
\end{align*}
%
Which simplifies to:
\begin{definition}[Generalised Termination Formula {\bf [GT]}]
\label{def:general-termination}
\begin{align*}
 \exists P_T, P_N. \forall x, x', y, \phi(P_T, x, x') \vee \psi(P_N, y)
\end{align*}
\end{definition}

Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other, as shown in Algorithm~\ref{alg:termination}.

\begin{algorithm}
 \caption{Bit-vector termination
 \label{alg:termination}}
 %\vspace{-1.5em}
 \begin{algorithmic}[1]
 \Statex
\Function{terminates}{$L$}
  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
  \Statex  \Comment{Formula {\bf[GT]}:}
  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
    \State \Return{$L$ terminates}
  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
    \State \Return{$L$ does not terminate}
  \EndIf
\EndFunction
 \end{algorithmic}
\end{algorithm}

Since $L$ either terminates or does not terminate, this formula is
satisfiable in the synthesis fragment.  A solution to the formula would
include witnesses $P_N$ and $P_T$, which are putative proofs of
non-termination and termination, respectively.  Exactly one of these will be
a genuine proof, so we can check first one and then the other.
